---
title: "Introduction to Geographic Data Science"
author: "Francisco Rowe ([`@fcorowe`](http://twitter.com/fcorowe))"
date: "`r Sys.Date()`"
output: tint::tintHtml
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tint)
# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
```

```{marginfigure}
[**Next**](01-gds.html)
```

# Description

This workshops offers an introduction to *Geographic Data Science*. It provides an introduction to fundamental concepts of geographic data science using a hands-on approach in *R*. It offers an overview of various types of spatial data, key challenges of working with these data, and some basic analytical techniques. 

# Structure

The workshop is structured as follows:

* [**What is geographic data science?**](01-gds.html)
* [**Spatial data**](02-spatial-data.html)  
* [**Spatial weights**](03-spatial_weights.html)  
* [**Spatial autocorrelation**]()  

# Philosophy

* Focus on **methods** and tools
  - General overview
  - Intuition
  - Very little math 
  - Opportunities to continue on your own
* Emphasis on the **application** and **use**
* Focus on **real-world** applications

# Resources

All this course material is available on Github and you can download it [**here**](https://github.com/fcorowe/udd_gds_course/archive/refs/heads/main.zip). Once you have download it, ensure it is in a safe place on your computer.

# Computational Environment

You need the most recent version of R and packages. These can be installed following the instructions provided in our [R installation guide](https://gdsl-ul.github.io/r_install/).

## Dependency list

The list of libraries that we will use is provided below. If you have followed the instructions provided in our [R installation guide](https://gdsl-ul.github.io/r_install/) and will be using Docker, you can relax and these libraries have already been installed for you. 

If you have `natively` installed *R* and *RStudio*, you need to ensure you have installed the list of libraries used in this book following the steps provided [here](https://gdsl-ul.github.io/r_install/otherWin.html#install-packages).

* `arm`
* `car`
* `corrplot`
* `FRK`
* `gghighlight`
* `ggplot2`
* `ggpubr`
* `ggmap`
* `GISTools`
* `gridExtra`
* `gstat`
* `jtools`
* `kableExtra`
* `knitr`
* `lme4`
* `lmtest`
* `lubridate`
* `MASS`
* `merTools`
* `plyr`
* `RColorBrewer`
* `rgdal`
* `sf`
* `sjPlot`
* `sp`
* `spgwr`
* `spatialreg`
* `spacetime`
* `stargazer`
* `tidyverse`
* `tint`
* `tmap`
* `viridis`


You can get the materials from this course as a [download](https://github.com/fcorowe/udd_gds_course/archive/refs/heads/main.zip) of a .zip file or by going directly to the [GitHub repository](https://github.com/fcorowe/udd_gds_course).





```{r eval=FALSE, include=FALSE}
file.edit(
  tint:::template_resources(
    'tint', '..', 'skeleton', 'skeleton.Rmd'
  )
)
```

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```