---
author: "<your-name>"
date: "<the-date>"
title: "Session 05"
output: html_document
---

## Class

Reopen your `vis-for-gds` project and save this `.Rmd` to the `reports` folder. When working through the materials from this session run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval=FALSE` on most code chunks, so when you knit the code chunk will not execute.

The technical element to this session involves analysing 2011 Census travel-to-work data.

### Setup

The code block below points to the packages you need to do the work discussed in the session. Uncomment the `intall_packages()` lines if you have not already downloaded `pct`.

```{r setup, echo=TRUE, eval=FALSE, message=FALSE}
library(tidyverse) # Bundle of packages for doing modern data analysis.
library(sf) # Spatial operations
library(here) # For navigating project directory

# remotes::install_github("ITSLeeds/pct")
library(pct)
```

### Import

Run the code block below to download and process the origin-destination Census data. This is a large datasets so this may take a little time to execute. Also import the London borough geometries -- real and regular grid.

```{r download-data, echo=TRUE, eval=FALSE, message=FALSE}
# Import OD by travel mode from 2011 Census.
od_pairs <- get_od(
  region = "london",
  type = "within",
  omit_intrazonal = FALSE,
  base_url = paste0("https://s3-eu-west-1.amazonaws.com/",
                    "statistics.digitalresources.jisc.ac.uk", "/dkan/files/FLOW/"),
  filename = "wu03ew_v2"
)

# Import .geojson file with geometry data for LondonSquared and real layout of London boroughs.
london_grid_real <- st_read("https://www.roger-beecham.com/datasets/london_grid_real.geojson")
```

Look-up table for London boroughs.

```{r borough-lookup, echo=TRUE, eval=FALSE, message=FALSE}
# Look-up of London boroughs.
london_las <- pct_regions_lookup %>% filter(region_name=="london") %>%
  pull(lad16nm)
```


Edges table:

```{r edges-table, echo=TRUE, eval=FALSE, message=FALSE}
edges <- od_pairs %>%
  # Filter only *within* London.
  filter(la_1 %in% london_las, la_2 %in% london_las) %>%
  group_by(la_1, la_2) %>%
  summarise(
    across(c(all:other), sum)
  ) %>%
  ungroup %>%
  transmute(
    o_bor=la_1, d_bor=la_2, all=all, public_transport=train+bus+light_rail,
    car=car_driver+car_passenger, active=bicycle+foot
  )
```

Nodes table

```{r nodes, echo=TRUE, eval=FALSE, message=FALSE}
# Summarise over destinations: commutes into- boroughs.
nodes_d <- od_pairs %>%
  # Filter only *within* London.
  filter(la_1 %in% london_las, la_2 %in% london_las) %>%
  group_by(la_2) %>%
  summarise(
    across(c(all:other), sum)
  ) %>%
  ungroup %>%
  rename(bor = la_2) %>%
  transmute(
    bor=bor, type="destination", all=all, public_transport=train+bus+light_rail,
    car=car_driver+car_passenger, active=bicycle+foot
  )

nodes_o <- od_pairs %>%
  # Filter only *within* London.
  filter(la_1 %in% london_las, la_2 %in% london_las) %>%
  group_by(la_1) %>%
  summarise(
    across(c(all:other), sum)
  ) %>%
  ungroup %>%
  rename(bor = la_1) %>%
  transmute(
    bor=bor, type="origin", all=all, public_transport=train+bus+light_rail,
    car=car_driver+car_passenger, active=bicycle+foot
  )

nodes  <- nodes_o %>% bind_rows(nodes_d)
```


## Homework

### Look-up tasks

#### For jobs filled in the City of London (CTY), from which borough does largest number of workers commute using public transport?

> *Enter your answer here*


#### Which borough in London has the largest *absolute* number living and working in that borough (and travelling to work by public transport)?

> *Enter your answer here*

#### For workers living in Barnet (BRN), what are the top 3 boroughs to which they commute?

> *Enter your answer here*


### Insights

Identify three isights that can be made around the geography of commuting patterns in London from reading any/each of the graphics. Briefly identify which graphic(s) led to the insight and how. Please do not spend too long over this activity -- the idea is to encourage you to make inferences from graphics, as you will do in your coursework.

#### Insight 1

> *Enter your answer here*

#### Insight 2

> *Enter your answer here*

#### Insight 3

> *Enter your answer here*
