---
author: "<your-name>"
date: "<the-date>"
title: "Session 04"
output: html_document
---

## Class

Reopen your `vis-for-gds` project and save this `.Rmd` to the `reports` folder. When working through the materials from this session run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval=FALSE` on most code chunks, so when you knit the code chunk will not execute.

The technical element to this session involves analysing [`stats19`](https://data.gov.uk/dataset/cb7ae6f0-4be6-4935-9277-47e5ce24a11f/road-safety-data) road crash data.

### Setup

The code block below points to the packages you need to do the work discussed in the session. Uncomment the `intall_packages()` lines if you have not already downloaded `stats19` and `trafficalmr`. Note that if you are having problems I ultimately make the main dataset used `ped_veh` available in a separate repo (see below "get-ped-veh" chunk).

```{r setup, echo=TRUE, eval=FALSE, message=FALSE}
library(tidyverse) # Bundle of packages for doing modern data analysis.
library(sf) # Spatial operations
library(here) # For navigating project directory
library(fst) # For reading/writing large files.
library(lubridate) # For working with dates.

# remotes::install_github("ropensci/stats19")
library(stats19)

# remotes::install_github("saferactive/trafficalmr")
library(trafficalmr)
```

### Import

Run the code block below to download and process the **crashes**, *vehicles** and **casualties** data. These are large datasets so this may take a little time to execute. **Note: if you are havung problems, skip to the third code block in this section, where the processed data can be loaded from an external repo.**

```{r download-data, echo=TRUE, eval=FALSE, message=FALSE}
# A vector containing years that we will collect data for.
years <- 2010:2019
crashes_all <- get_stats19(year = years, type = "ac")
casualties_all <- get_stats19(year = years, type = "cas")
vehicles_all <-  get_stats19(year = years, type = "veh")

# Use trafficalmr package for recoding.
casualties_all <- casualties_all %>% mutate(casualty_type=tc_recode_casualties(casualty_type))
vehicles_all <- vehicles_all %>% mutate(vehicle_type=tc_recode_vehicle_type(vehicle_type))
crashes_all <- crashes_all %>% mutate(speed_limit=tc_recode_speeds_uk(speed_limit))

# Large files, so write out to `.fst`.
write_fst(crashes_all, here("data", "crashes_all.fst"))
write_fst(casualties_all, here("data", "casualties_all.fst"))
write_fst(vehicles_all, here("data", "vehicles_all.fst"))

# And read in as`.fst`.
vehicles_all <- read_fst(here("data", "vehicles_all.fst"))
casualties_all <- read_fst(here("data", "casualties_all.fst"))
crashes_all <- read_fst(here("data", "crashes_all.fst"))
```

Run the code block below to generate the derived dataset of pedestrian casualties used in the class session and homework.

```{r subset-pedestrians, echo=TRUE, eval=FALSE, message=FALSE}
# Data frame of all pedestrian casualties.
ped_veh <- crashes_all %>%
  left_join(casualties_all %>%
              select(accident_index, age_of_casualty,sex_of_casualty, casualty_type, casualty_imd_decile, casualty_severity,
                     pedestrian_location, pedestrian_movement, casualty_reference))  %>%
  filter(casualty_type=="Pedestrian")

# Ordered factor of vehicle types.
vehicles_all %>% select(vehicle_type) %>% distinct() %>% pull()
veh_orders <- c("Bicycle", "Motorcycle","Car", "Taxi", "Other", "Van", "Bus", "HGV")
# Filtering join to filter all vehicles involved in pedestrian crashes.
# Then identify largest vehicle involved in each crash and filter on these.
# Then identify "first" vehicle_reference involved in each crash.
temp_ped_vehicles <- vehicles_all %>%
  semi_join(ped_veh) %>%
  mutate(vehicle_type=factor(vehicle_type, levels = veh_orders, ordered = TRUE)) %>%
  group_by(accident_index) %>%
  mutate(largest_vehicle = max(vehicle_type)) %>%
  filter(vehicle_type==largest_vehicle) %>%
  mutate(unique_vehicle=min(vehicle_reference)) %>%
  filter(vehicle_reference==min(vehicle_reference))

ped_veh <- ped_veh %>% left_join(temp_ped_vehicles)
# Remove temporary/staging data frame.
rm(temp_ped_vehicles)
```

**All**: You may have had problems due to the fact that we are working with a large dataset, so I have made `ped_veh` available on a personal repository. Run the block below to download.

```{r get-ped-veh, echo=TRUE, eval=FALSE, message=FALSE}
# Dataset of pedestrian-vehicle crashes.
url <- "https://www.roger-beecham.com/datasets/ped_veh.fst"
download.file(url, here("data", "ped_veh.fst"))
ped_veh <- fst::read_fst(here("data", "ped_veh.fst"))
```

Run the code block below  to recode IMD to quintiles.


```{r get-imd, echo=TRUE, eval=FALSE, message=FALSE}
# Recode imd into quintiles.
ped_veh <- ped_veh %>%
  mutate(
    driver_imd_quintile=case_when(
      driver_imd_decile == "Most deprived 10%" ~ "1 most deprived",
      driver_imd_decile == "More deprived 10-20%" ~ "1 most deprived",
      driver_imd_decile == "More deprived 20-30%" ~ "2 more deprived",
      driver_imd_decile == "More deprived 30-40%" ~ "2 more deprived",
      driver_imd_decile == "More deprived 40-50%" ~ "3 mid deprived",
      driver_imd_decile == "Less deprived 40-50%" ~ "3 mid deprived",
      driver_imd_decile == "Less deprived 30-40%" ~ "4 less deprived",
      driver_imd_decile == "Less deprived 20-30%" ~ "4 less deprived",
      driver_imd_decile == "Less deprived 10-20%" ~ "5 least deprived",
      driver_imd_decile == "Least deprived 10%" ~ "5 least deprived",
      TRUE ~ driver_imd_decile),
    casualty_imd_quintile=case_when(
      casualty_imd_decile == "Most deprived 10%" ~ "1 most deprived",
      casualty_imd_decile == "More deprived 10-20%" ~ "1 most deprived",
      casualty_imd_decile == "More deprived 20-30%" ~ "2 more deprived",
      casualty_imd_decile == "More deprived 30-40%" ~ "2 more deprived",
      casualty_imd_decile == "More deprived 40-50%" ~ "3 mid deprived",
      casualty_imd_decile == "Less deprived 40-50%" ~ "3 mid deprived",
      casualty_imd_decile == "Less deprived 30-40%" ~ "4 less deprived",
      casualty_imd_decile == "Less deprived 20-30%" ~ "4 less deprived",
      casualty_imd_decile == "Less deprived 10-20%" ~ "5 least deprived",
      casualty_imd_decile == "Least deprived 10%" ~ "5 least deprived",
      TRUE ~ casualty_imd_decile)
  )

# IMD data from:   https://geoportal.statistics.gov.uk/datasets/3db665d50b1441bc82bb1fee74ccc95a_0
imd <- read_csv("https://opendata.arcgis.com/datasets/3db665d50b1441bc82bb1fee74ccc95a_0.csv")
imd <- imd %>% mutate(
  quintile=ntile(IMD19,5),
  crash_quintile=case_when(
    quintile == 1 ~ "1 most deprived",
    quintile == 2 ~ "2 more deprived",
    quintile == 3 ~ "3 mid deprived",
    quintile == 4 ~ "4 less deprived",
    quintile == 5 ~ "5 least deprived"
  )
)

ped_veh <- ped_veh %>%
  # Crashes in Wales are excluded as English IMD
  inner_join(imd %>% select(LSOA11CD, crash_quintile), by=c("lsoa_of_accident_location"="LSOA11CD"))
```

## Homework

### Visualization design challenge

Uncomment the code block below (remove the "#" and insert the path to you figure as an argument to the `include_graphics()` function.

<!-- ![](<insert-path-to-figure-here>) -->

# ```{r vis-design-challenge, echo=FALSE, fig.cap="<insert figure caption here>", fig.align="left",  out.width="100%"}
# include_graphics("<path-to-figure>", error = FALSE)
# ```

```{r design-challenge-code, echo=TRUE, eval=FALSE}
#######################
# Enter your code in the chunk provided.
######################
```


| Data item      | Measurement level | Visual mark      | Visual channel  |
|----------------|-------------------|------------------|-----------------|
| `location`     | <enter here>      |  <enter here>    | <enter here>    |
| `...`          |`...`              |  `...`           | `...`           |
| `...`          |`...`              |  `...`           | `...`           |
| `...`          |`...`              |  `...`           | `...`           |
| `...`          |`...`              |  `...`           | `...`           |
| `...`          |`...`              |  `...`           | `...`           |
