---
author: "<your-name>"
date: "<the-date>"
title: "Session 06"
output: html_document
---

## Class

Reopen your `vis-for-gds` project and save this `.Rmd` to the `reports` folder. When working through the materials from this session run each code chunk separately by clicking the small green arrow beside each code chunk. By default I have set `eval=FALSE` on most code chunks, so when you knit the code chunk will not execute.

The technical element to this session involves analysing estimated vote shares for Leave in the UK's 2016 Referendum on EU membership.

### Setup

The code block below points to the packages you need to do the work discussed in the session. Uncomment the `intall_packages()` lines if you have not already downloaded `tidymodels`.

```{r setup, echo=TRUE, eval=FALSE, message=FALSE}
library(tidyverse) # Bundle of packages for doing modern data analysis.
library(sf) # Spatial operations
library(here) # For navigating project directory

library(parlitools) # Contains required data.
# install.packages("tidymodels") # Uncomment for download.
library(tidymodels) # For working with model outputs.
```

### Import

Run the code block below to download and process the data required for this session.

```{r collect-data, echo=TRUE, eval=FALSE, message=FALSE}
# Load constituency boundaries, simplified using mapshapr
url <- "https://www.roger-beecham.com/datasets/cons_outline.geojson"
cons_outline <- st_read(url, crs=27700)

# Collect explanatory variables (from census_11 table in parlitools)
explanatory <- census_11 %>%
  transmute(
    ons_const_id=ons_const_id, constituency_name=constituency_name, region=region,
    population=population, population_density=population_density,
    younger=age_20_to_24+age_25_to_29+age_30_to_44,
    own_home=house_owned,
    no_car=cars_none, white=ethnicity_white_british+ethnicity_white_irish,eu_born=born_other_eu, christian,
    professional=nssechigher_manager+nssechigher_professional, degree=qual_level_4,
    not_good_health=health_fair+health_bad+health_very_bad, heavy_industry=industry_manufacturing+industry_transport
    )

  # Collect outcome variable (from leave_votes_west table in parlitools)
  outcome <- leave_votes_west %>%
    select(ons_const_id, constituency_name, leave=figure_to_use) %>%
    inner_join(explanatory %>% select(ons_const_id, region)) %>%
    mutate(gb_leave=.519, resid_unform=leave-gb_leave)

# Z-score transform explanatory variables.
explanatory_z_scores <- explanatory %>%
  mutate(
    across(
      .cols=c(younger:heavy_industry), .funs=~(.x-mean(.x))/sd(.x)
    )
  )

# Generate data frame containing data for models.
data_for_models <- explanatory_z_scores %>%
  inner_join(outcome %>% select(ons_const_id, leave, region))
```

## Homework


### Task 1: Calculate and plot

```{r calculate-correlations, echo=TRUE, eval=FALSE, message=FALSE}
# Place your code for calculating correlation coefficients by explanatory
# variable here. Note that you will then want to extract a vector of ordered
# variable names as an object.

```


```{r plot-scatters, echo=TRUE, eval=FALSE, message=FALSE}
# Place your code for generating scatterplots here.
```

### Task 2: Explore, model and evaluate


```{r map-residuals, echo=TRUE, eval=FALSE}
#######################
# Enter your code in the chunk provided.
######################
```


#### Insight 1

> *Enter your answer here*

#### Insight 2

> *Enter your answer here*

#### Insight 3

> *Enter your answer here*
