---
title: "Model building 1"
linktitle: "6: Model building 1"
date: "2022-06-13"
class_date: "2022-06-13"
citeproc: true
bibliography: ../../static/bib/references.bib
csl: ../../static/bib/chicago-author-date.csl
output:
  blogdown::html_page:
    template: ../../pandoc/toc-title_html.template
    toc: true
menu:
  homework:
    parent: Homework
    weight: 1
type: docs
weight: 1
editor_options:
  chunk_output_type: console
---

```{r setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(knitr)
```

## Session Outcomes

By the end of this homework session you should be able to:

{{% callout objective %}}
- [x] Calculate grouped statistics and use those statistics for ordering plot outputs.
- [x] Extract parameters from linear regression models and explore their structure graphically.
- [x] Make use of functional programming techniques for working over datasets.
{{% /callout %}}

## Introduction

This homework requires you to apply the concepts and skills developed in the class session on [model building 1](/class/06-class/). Do complete the task and be sure to save your work as it will be submitted as part of your portfolio for Assignment 1.

## Task 1: Calculate and plot

Recreate the [scatterplots in the class materials](/class/06-class/#quantifying-and-exploring-co-variation-1) displaying the Leave against candidate explanatory variable -- faceted by each explanatory variable and ordered by correlation coefficient.


You will need to use the `data_for_models` data frame. First, calculate correlation coefficients for each explanatory variable, then generate a vector of variable names, stored in an object, ordered according to correlation coefficient.  To help you out, I've provided a template of `dplyr` commands that you will want to use and some indicative output. In order to summarise over each explanatory variable in a `group_by()`, you first need to `pivot_longer()` so that all explanatory variables are represented in a single column. Of course you are welcome to take a different approach.

```{r calculate-correlations, echo=TRUE, eval=FALSE, message=FALSE}
# Place your code for calculating correlation coefficients by explanatory
# variable here.

expl_cors <- data_for_models %>%
  pivot_longer(...) %>%
  group_by(...) %>%
  summarise(...) %>%
  arrange(...)

expl_cors
# # A tibble: 10 x 2
#    expl_var           cor
#    <chr>            <dbl>
#  1 degree          -0.772
#  2 professional    -0.565
#  3 younger         -0.538
#  4 eu_born         -0.483
#  5 no_car          -0.396
#  6 white            0.411
#  7 own_home         0.430
#  8 christian        0.487
#  9 not_good_health  0.562
# 10 heavy_industry   0.710

# You will then want to then extract a vector of ordered
# variable names as an object.

expl_cors_order
# [1] "degree"          "professional"    "younger"         "eu_born"
# [5] "no_car"          "white"           "own_home"        "christian"
# [9] "not_good_health" "heavy_industry"

```

Now generate the scatterplots with Leave against each explanatory variable, faceted by explanatory variable and with the facets ordered according to correlation coefficient.

```{r scatters-code, echo=TRUE, eval=FALSE}
#######################
# Enter your code in the chunk provided.
######################
```

* [Save your plot](https://r4ds.had.co.nz/graphics-for-communication.html#saving-your-plots) as a `.png` file and insert the file as a Figure into your `.Rmd` document ([guidance here](https://rmd4sci.njtierney.com/figures-tables-captions-.html)).


## Task 2: Explore, model and evaluate

A variable that is intuitively relevant to this study, but that behaves counter to expectation is *EU-born* -- the proportion of residents in a constituency born outside of the UK but within the EU. In the [technical](/class/06-class/#techniques) element to the session, you generated a data frame of model outputs regressing each candidate explanatory variable on Leave. This is stored in `single_model_fits`. The second set of [scatterplots](class/06-class/#modelling-for-co-variation-1) use information contained in `single_model_fits` to colour observations according to their residuals from each of these model objects. The scatterplot for *EU-born* is interesting -- some observations with very large negative coefficients and some with very large positive coefficients.

Extract the residuals from the *EU-born* model object contained in `single_model_fits`. Then map it in a similar way to the [map-lineup](/class/06-class/#exploring-bias-1) -- that is, a red-blue colour scheme centred on 0 (not the line-up itself).

```{r map-residuals, echo=TRUE, eval=FALSE}
#######################
# Enter your code in the chunk provided.
######################
```
* [Save your plot](https://r4ds.had.co.nz/graphics-for-communication.html#saving-your-plots) as a `.png` file and insert the file as a Figure into your `.Rmd` document ([guidance here](https://rmd4sci.njtierney.com/figures-tables-captions-.html)).

What do you notice about the geography of these residuals? Make a few short observations in the template provided.

* Insight 1
  + *Enter your answer in the template*
* Insight 2
  + *Enter your answer in the template*
* Insight 3
  + *Enter your answer in the template*



## References
